Generating Empirically Optimized Composed Matrix Kernels from MATLAB Prototypes

### 2008-12-18 doc/papers/iccs09/paper.tex
removed formatting hacks, cut down to 10 pages again.
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@402 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

% formatting cheats
\renewcommand{\baselinestretch}{0.92}  
\renewcommand\floatpagefraction{.9} 
\renewcommand\topfraction{.9} 
\renewcommand\bottomfraction{.9} 
\renewcommand\textfraction{.1} 
\setcounter{totalnumber}{50} 
\setcounter{topnumber}{50} 
\setcounter{bottomnumber}{50} 
\setlength{\textwidth}{12.5cm} 
\setlength{\textheight}{19.5cm}

@@

\author{Boyana Norris\thanks{The work of this author was supported by the Office of Advanced Scientific Computing Research, Office of Science, U.S. Dept. of Energy, under Contract DE-AC02-06CH11357.}\inst{1}
  \and Albert Hartono\inst{2}
  \and Elizabeth Jessup\thanks{The work of this author was supported by the
  National Science Foundation under grants no. CCF-0430646 and CCF-830458.}\inst{3}
  \and Jeremy Siek\inst{3}
%
\authorrunning{Norris, Hartono, Jessup, Siek}
%
\institute{Argonne National Laboratory \\
  \email{norris@mcs.anl.gov}
  \and Ohio State University \\
  \email{hartonoa@cse.ohio-state.edu}
  \and University of Colorado at Boulder \\
  \email{\{elisabeth.jessup,jeremy.siek\}@colorado.edu}}}

@@

The development of high-performance numerical codes is challenging because
performance is determined by complex interactions between the algorithm, data
structure, programming language, compiler, and computer
architecture. Scientists seeking high performance are thus required to master
advanced concepts in computer science and carry out intricate programming
tasks in addition to managing the scientific content of their work.  They
must either invest substantial time in tuning their software or accept low
performance. In either case, the productivity of the scientist degrades.

Historically, the research community has pursued two separate paths toward
the goal of making software run at near-peak levels.  The first path builds
on research into compilers and their associated technologies.  One of the
main goals of compilation research is to take an arbitrary code as input and
produce optimal code as output for a given language and hardware platform.
The success of this approach has been limited by a number of factors: (i)
optimal mappings between the computation graph and the hardware are expensive
(often NP-complete) to compute; (ii) potentially useful information that
could aid optimization cannot be represented in general-purpose languages
such as C and Fortran; and (iii) user control of compiler optimizations is
limited and varies from compiler to compiler, and (iv) apart from differences
in execution time, it is difficult for users to evaluate the effectiveness of
different compiler optimizations.

The second path to performance optimization is to identify kernel
routines that dominate the execution time of a wide variety of
applications.  
%When such kernels can be identified and agreed upon by the
%members of the community, programmers with the required level of technical
%knowledge can concentrate on producing these optimized kernel libraries for
%architectures of interest. 
An example is the high-performance BLAS libraries produced by a combination
of hardware vendors, independent software vendors, and researchers.
Developers who write their codes calling these routines can achieve high
performance across all supported architectures. The efforts in
high-performance library development aptly demonstrate, however, that the
research into compiler technologies has not been fully successful.  If
compilers could take the relatively simple code such as a dense matrix-matrix
product and produce optimal code, this second approach would be unnecessary.
Like compiler optimization, the kernel library-oriented approach has
significant limitations.  Support for a variety of architectures is the most
serious problem because the optimizations necessary to achieve near-peak
performance are by nature not portable, requiring manual update or
reimplementation for each new architecture.

This paper describes a combination of the two approaches designed to overcome
some of their shortcomings. We take a multipronged approach
toward the development of software infrastructure for \emph{generating}
automatically tuned libraries for matrix algebra computations.  We briefly
overview relevant prior and current research efforts in
Section~\ref{sec:background}. In Section~\ref{sec:approach} we describe our
MATLAB~\cite{matlab:webpage} to C compiler and the empirical performance tuning system Orio and its
use in conjunction with the PLuTo tool suite to generate and empirically
evaluate many tuned versions of the C code generated by the MATLAB compiler.
In Section~\ref{sec:experiments}, we provide performance results showing that
the code generated by using our system significantly outperforms not only the
original simple C code but also code based on source BLAS, ATLAS-optimized
BLAS, and Intel MKL routines.
%Text in intro repeats abstract some, but that might not be a big deal for a conf paper.

@@

Existing optimizing MATLAB compilers, such as the MaJIC MATLAB compiler~\cite{MaJIC},
include limited local optimizations for matrix expressions, but do not
perform optimizations such as loop fusion across multiple operations as we do
with the tools described in this paper. The telescoping languages
%~\cite{telescopingurl,teleoverview,Ken99} 
project~\cite{telescopingurl} uses techniques such as
strength reduction, vectorization, and procedure specialization to optimize
MATLAB scripts, but do not generate reusable optimized linear algebra
routines as described in this paper.

@@

Codes based on matrix algebra are generally constructed as a sequence of
calls to the Basic Linear Algebra Subprograms
(BLAS)
%~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} 
and similar
sparse matrix libraries~\cite{Saad:fr,George:1981uq}. Writing programs in
this way promotes readability and maintainability but can be costly in terms
of memory efficiency. Specifically, the retrieval of a large order matrix at
each routine call can have a profound effect on performance even when highly
tuned implementations of the BLAS (e.g.,~\cite{Goto:2006fk})
%(e.g.~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk})
are used.

The authors and other researchers have observed that it can be much more
efficient to call a single specialized routine that performs multiple
operations than to make successive calls to separate BLAS routines (e.g., see
~\cite{gropp01,Vuduc:2003kl}.
%~\cite{Ashby:uq,Blackford:2002vn,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
Single routines that carry out more than one linear algebra operation are
known as \emph{composed} BLAS. As an example, consider the pair of
matrix-vector products $q = Ap$, $s = A^Tr$, where $A$ is a matrix and $p$,
$q$, $r$, and $s$ are vectors, that comprise the computational bottlenecks of
the BiConjugate Gradient Method~\cite{Barrett:1994kx,Saad:2003fk} and of the
GEMVER kernel examined in Section~\ref{sec:matlab}.  These two operations can
be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can
be rewritten as a composed BLAS consisting of the doubly nested loop
encompassing both matrix-vector products. In the former case, the matrix
$A$ is accessed twice, while in the latter it is accessed only once. Our
preliminary results in prior research and in this paper indicate that 
loop fusion leads to a routine that delivers significantly better performance
than does a pair of calls to the best optimized BLAS {\tt GEMV} routines for
large matrix orders.  Composed routines are the focus of the work presented
here, but much of what we discuss generalizes to a much broader array of
computations.

To generate highly optimized code from a MATLAB prototype of the composed BLAS operation, we follow a three-step approach, illustrated in Figure~\ref{fig:process}.  To begin,
we have developed a compiler that converts a MATLAB script into simple C code~\cite{Siek}.
After generating the C code from the high-level MATLAB prototype, 
%
\begin{wrapfigure}{r}{3in}
\vspace{-.2in}
\centering
\includegraphics[width=3in]{figures/process.png}
\caption{The code generation and tuning process.}
\label{fig:process}
\vspace{-.2in}
\end{wrapfigure}
%
we (optionally) use the source-to-source automatic parallelization tool PLuTo~\cite{Pluto} %developed at Ohio State University
to optimize for coarse-grained parallelism and locality simultaneously. Using
the results of the PLuTo analysis, we insert annotations into the C code,
which are then processed by our extensible annotation system Orio to generate
many tuned versions of the same operation using different optimization
parameters. Orio then performs an empirical search to select the best among
the multiple optimized code variants.

@@

%\vspace{-.2in}
\subsection{A MATLAB Compiler}
\label{sec:matlab}
%\input{matlab}
In this section we describe the design and operation of the MATLAB to C compiler we have developed~\cite{Siek}.    Figure~\ref{fig:compiler} gives an overview of the compilation process. 
%
\begin{wrapfigure}{r}{2in}
\vspace{-.2in}
\centering
\includegraphics[width=2in]{figures/compile.png}
\caption{The MATLAB compiler.}
\label{fig:compiler}
\vspace{-.2in}
\end{wrapfigure}
The MATLAB kernel specification is parsed into a high-level intermediate
representation in the form of a dataflow graph.  This dataflow graph is then
iteratively processed until all of the implementation choices have been made.
The compilation process consists of three phases: analysis, refinement, and
optimization, which are together iterated until all of the implementation
decisions have been made.  The graph is then translated into C code.
%\subsubsection{The Dataflow Graph.}
%
An example dataflow graph for the GEMVER kernel, defined below, is shown
in Figure~\ref{fig:gemver-dataflow}.
Each node represents a parameter of the kernel or an operation.  The arrows
indicate the flow of data. At first the graph specifies what operations are
to be performed but does not contain any implementation details. The symbol
$\times$ in the depicted graph stands variously for outer product,
matrix-vector multiplication, and scalar-vector multiplication, and it does
not yet specify, for example, whether the outer products compute a row or a
column at a time of the result matrix.
\begin{figure}[ht]
\begin{minipage}{.3\textwidth}
\begin{align*}
  A &\gets A + u_1 v_1^T + u_2 v_2^T \\[-0.5ex]
  x &\gets \beta A^T y + z \\[-0.5ex]
  w &\gets \alpha A x
\end{align*} 
\end{minipage}
\begin{minipage}{.6\textwidth}
\includegraphics[width=3in]{figures/gemver-dataflow.png}
\end{minipage}
\caption{Dataflow graph (right) for the GEMVER kernel (left).}
\label{fig:gemver-dataflow}
\vspace{-.2in}
\end{figure}

@@

%Once
%we add backtracking we will push each alternative onto the stack to be
%explored later. Once the choice is made, the kind and storage format labels
%for the operation node itself and for the input nodes are updated with the
%labels specified in the linear algebra database for the chosen algorithm. The
%algorithm then goes on to find the next node with the fewest matching
%implementations and repeats.  In this phase, only the name of the algorithm
%is assigned to the operation node.  The details of the implementation are not
%specified until the refinement phase.

The refinement phase resolves the implementation for each operation node in
the graph into a subgraph defining the details of the chosen algorithm.  Each
subgraph is an abstract representation of the loop that implements the given
operation that also contains an iteration strategy for traversing the
elements of the matrix or vector.

@@

In the optimization step, we apply conditional rewrite rules to optimize the
dataflow graph, for example merging two subgraphs when they share a common
operand.
%but are not dependent on one another.
%  One example rewrite rule is
%\emph{Merge Operand Sharing Subgraphs}: if two subgraphs share a common
%operand but do not depend on one another, and if the iteration strategies of
%the two subgraphs are compatible, then merge the subgraphs into a single
%subgraph.  
This rule is responsible for fusing the loops of the two
matrix-vector products in the GEMVER kernel.  
%Another example rule is
%\emph{Merge Dependent Subgraphs}: if one subgraph depends on another
%subgraph, and the iteration strategies of the two subgraphs are compatible,
%then merge them into a single subgraph.  This rule, for example, is
%responsible for fusing the loops of the two outer products in the GEMVER
%kernel.

%The choice of which optimizations to apply to various parts of the graph
%depends on the characteristics of the kernel, the algorithm, the matrix order
%and kind, and architectural features.

%The following is an instructive
%example: $y \gets A^T A (x + z + w)$.  Assume that the compiler has already
%merged the two vector additions into one subgraph, call it $S_1$, and the two
%matrix products into another subgraph, call it $S_2$. The question then is
%whether it is profitable to apply the Merge Dependent Subgraphs rule one more
%time to merge $S_1$ and $S_2$. If the three vectors plus a row of $A$ do not
%fit in cache, then the merger will not be profitable because it will cause
%each row of $A$ to fall out of cache before the row gets used a second time.
%On the other hand, if all three vectors plus a row of $A$ do it in cache,
%then $S_1$ and $S_2$ should be merged.  In general, we need a way to search
%through the many different combinations of optimization choices and choose a
%combination that minimizes memory traffic.

%\subsubsection{Translate the Dataflow Graph to C.}

The final step performed by the MATLAB compiler when the graph cannot be
refined further is the generation of C code. The generator outputs a C loop
for each subgraph based on a topological sort of the graph.

@@

\begin{figure}[tb]
\vspace{-.3in}
\begin{center}
\includegraphics[width=.65\textwidth]{figures/orio.png}
\end{center}
\vspace{-.25in}
\caption{Overview of the Orio empirical tuning process.}
\label{fig:orio}
\vspace{-.2in}
\end{figure}

@@

\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.45\textwidth}
\scriptsize
\begin{verbatim}
spec vadd_tune_spec {
 def build { 
  arg build_command = 'mpixlc -O3 -qstrict -lm';
  arg batch_command = 'qsub -n 64 -t 10';
  arg status_command = 'qstat';
  arg num_procs = 64;
 }
 def performance_params { 
  param UF[] = range(1,32);
  param PAR[] = [True, False];
 } 
 def input_params { 
  param N = [10,100,1000,10**4,10**5,10**6,10**7];
 } 
 def input_vars {
  decl int n = N;
  decl double y[N] = 0;
  decl double x1[N] = random;   
  decl double x2[N] = random;   
  decl double x3[N] = random;   
 }
 def search {
  arg algorithm = 'Exhaustive';
 }
}
\end{verbatim}
\end{minipage}\\
\end{tabular}
\vspace{-.1in}
\caption{Orio example: Annotated C source code (left) and tuning specification excerpt for the Blue Gene/P (right).}
\label{fig:orio-example}
\vspace{-.2in}
\end{figure}

Figure~\ref{fig:orio-example} shows an annotation example used by Orio
to empirically optimize VADD operation on Blue Gene/P. The annotations
contain performance hints that instruct Orio to perform memory
alignment optimization, loop unrolling, and multicore parallelization
(using OpenMP). In addition to these simple optimizations, Orio
supports other transformations such as loop blocking, loop
permutation, scalar replacement, array copy optimization, and other
architecture-dependent optimizations (e.g., SIMDization for Intel and
Blue Gene/P architectures). The right-hand side of
Figure~\ref{fig:orio-example} shows separate tuning specifications
used for building and running executable tests, including performance
parameter values, execution environment details, input variable
information, and the search algorithm. Orio also supports parallel
search when parallel resources are available. In this example, the
parallel Orio driver simultaneously executes 64 code variants in the
same parallel job. The commands used by Orio to submit a parallel job
and to query its status are also specified in the tuning
specifications. At present, users must create the tuning
specifications manually. When Orio is used in conjunction with
compiler tools, such as the MATLAB compiler described in this paper,
it should eventually be possible to automatically generate the tuning
specifications.

\vspace{-0.1in}

@@

Table~\ref{tbl:blas-ops} lists the composed BLAS operations used in
our experiments, along with their input and output variables. Vectors
are typeset in lowercase with an overhead arrow. Scalars and matrices
are represented as lowercase and uppercase letters, respectively. A
regular uppercase denotes a row matrix, whereas a bold uppercase
symbolizes a column matrix. The extended MATLAB code that corresponds
to each operation can be seen in the last column of
Table~\ref{tbl:blas-ops}.

\begin{table}[htb] 
\vspace{-.1in} 
\caption{Composed BLAS operations used in our experiments.}
\label{tbl:blas-ops}
\small
\centering 
%\vspace{-.1in} 
\begin{tabular}{|c|c|c|c|} 
\hline 
Name & Input & Output & C code from MATLAB compiler \\ 
\hline 
\hline 
%& & & \\
VADD & $\overrightarrow{w}$,$\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$ &
$\overrightarrow{x} = \overrightarrow{w} + \overrightarrow{y} + \overrightarrow{z}$ \\ [1ex]
\hline 
%& & & \\
ATAX & $A$,$\overrightarrow{x}$ & $\overrightarrow{y}$ & 
$\overrightarrow{y} = A' * (A * \overrightarrow{x})$ \\ [1ex]
\hline 
%& & & \\
 & ${\bf A}$,$a$,$b$, &  & ${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
GEMVER & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & ${\bf B}$,  & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
& $\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\ [1ex]
\hline 
& & & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\ [1ex]
\hline 
\end{tabular} 
%\vspace{-.2in} 
\end{table}

@@

\end{tabular}
}
\caption{Performance results for different implementations of several composed BLAS operations.}
\label{fig:results}
\end{figure}

The performance results of tuning the VADD operation are given in
Figure~\ref{fig:results}(a). The ``Base'' label designates the simple
loop C implementation generated by the MATLAB compiler, which was
simply compiled with the options mentioned earlier and executed. We
tested the performance of the IBM ESSL library and Goto BLAS for the
two calls to \texttt{daxpy} used in the BLAS version of the code.
Finally, we tuned the simple C loop version using Orio, with the
performance annotations previously shown in
Figure~\ref{fig:orio-example}. In this experiment, we measured the
performance for both the sequential and parallel scenarios. We observe
that even for a very simple operation, such as the vector addition
above, the compiler alone is unable to obtain the same level of
performance as the empirically tuned versions. Furthermore, as
expected, the BLAS performance was worse than that of the single loop
implementation, however, selecting the correct compiler options was
crucial for achieving good performance of the base version.

@@

The performance results of ATAX are shown in
Figure~\ref{fig:results}(b). The Orio-tuned version that incorporates
PLuTo-generated loop fusion optimizations and Orio parallelization
directives achieves the best performance for most problem sizes,
outperforming the Intel MKL version by a factor of 2 to 5.7 and the
compiler-optimized C version by a factor of 4 to 7. The optimizations
performed by both Orio versions include scalar replacement,
vectorization, and loop unroll/jam.

>>>>

% formatting cheats
%\renewcommand{\baselinestretch}{0.92}  
%\renewcommand\floatpagefraction{.9} 
%\renewcommand\topfraction{.9} 
%\renewcommand\bottomfraction{.9} 
%\renewcommand\textfraction{.1} 
%\setcounter{totalnumber}{50} 
%\setcounter{topnumber}{50} 
%\setcounter{bottomnumber}{50} 
%\setlength{\textwidth}{12.5cm} 
%\setlength{\textheight}{19.5cm}

@@

\author{Boyana Norris\thanks{The work of this author was supported by the Office of Advanced Scientific Computing Research, Office of Science, U.S. Dept. of Energy, under Contract DE-AC02-06CH11357.}\inst{1}
  \and Albert Hartono\inst{2}
  \and Elizabeth Jessup\thanks{The work of this author was supported by the
  National Science Foundation under grants no. CCF-0430646 and CCF-830458.}\inst{3}
  \and Jeremy Siek\inst{3}
%
\authorrunning{Norris, Hartono, Jessup, Siek}
%
\institute{Argonne National Laboratory \\
  \email{norris@mcs.anl.gov}
  \and Ohio State University \\
  \email{hartonoa@cse.ohio-state.edu}
  \and University of Colorado at Boulder \\
  \email{\{elizabeth.jessup,jeremy.siek\}@colorado.edu}}}

@@

The development of high-performance numerical codes is challenging because
performance is determined by complex interactions among the algorithm, data
structure, programming language, compiler, and computer
architecture. Scientists seeking high performance are thus required to master
advanced concepts in computer science and carry out intricate programming
tasks in addition to managing the scientific content of their work.  They
must either invest substantial time in tuning their software or accept low
performance. In either case, the productivity of the scientists degrades.

Historically, the research community has pursued two separate paths toward
the goal of making software run at near-peak levels.  The first path builds
on research into compilers and their associated technologies.  One of the
main goals of compilation research is to take an arbitrary code as input and
produce optimal code as output for a given language and hardware platform.
The success of this approach has been limited by a number of factors: (i)
optimal mappings between the computation graph and the hardware are expensive
(often NP-complete) to compute; (ii) potentially useful information that
could aid optimization cannot be represented in general-purpose languages
such as C and Fortran; and (iii) user control of compiler optimizations is
limited and varies from compiler to compiler, and (iv) apart from differences
in execution time, it is difficult to evaluate the effectiveness of
different compiler optimizations.

When compilers alone cannot achieve the desired performance, 
another path to performance optimization is to identify kernel
routines that dominate the execution time of a wide variety of
applications.  
%When such kernels can be identified and agreed upon by the
%members of the community, programmers with the required level of technical
%knowledge can concentrate on producing these optimized kernel libraries for
%architectures of interest. 
An example is the high-performance Basic Linear Algebra Subprograms 
(BLAS) libraries~\cite{BLAS} produced by a combination
of hardware vendors, independent software vendors, and researchers.
Developers who write their codes calling these routines can achieve high
performance across all supported architectures, but are also subject to 
the limitations of the library (e.g., portability and the types of operations available).
%The efforts in
%high-performance library development aptly demonstrate, however, that the
%research into compiler technologies has not been fully successful.  If
%compilers could take the relatively simple code such as a dense matrix-matrix
%product and produce optimal code, this second approach would be unnecessary.
%Like compiler optimization, the kernel library-oriented approach has
%significant limitations.  Support for a variety of architectures is the most
%serious problem because the optimizations necessary to achieve near-peak
%performance are by nature not portable, requiring manual update or
%reimplementation for each new architecture.

This paper describes a combination of the two approaches designed to overcome
some of their shortcomings. We describe our initial efforts toward the
development of software infrastructure for \emph{generating} automatically
tuned libraries for matrix algebra computations.  In
Section~\ref{sec:background} we briefly discuss relevant prior and current
research efforts. In Section~\ref{sec:approach} we describe our
MATLAB-to-C compiler and the empirical performance
tuning system Orio and its use in conjunction with the PLuTo tool suite to
generate and empirically evaluate many tuned versions of the C code generated
by the MATLAB compiler.  In Section~\ref{sec:experiments} we provide
performance results on two architectures. We conclude with a summary and
discussion future work in Section~\ref{sec:conclusion}.

%showing that the code generated by using our system
%significantly outperforms not only the original simple C code but also code
%based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
%Text in intro repeats abstract some, but that might not be a big deal for a conf paper.

@@

Existing optimizing MATLAB~\cite{matlab:webpage} compilers, such as the MaJIC
MATLAB compiler~\cite{MaJIC}, include limited local optimizations for matrix
expressions, but do not perform optimizations such as loop fusion across
multiple operations as we do with the tools described in this paper. The
telescoping languages
%~\cite{telescopingurl,teleoverview,Ken99} 
project~\cite{telescopingurl} uses techniques such as
strength reduction, vectorization, and procedure specialization to optimize
MATLAB scripts, but do not generate reusable optimized linear algebra
routines as described in this paper.

@@

Codes based on matrix algebra are generally constructed as a sequence of
calls to the Basic Linear Algebra Subprograms
(BLAS)
%~\cite{Dongarra:1988uq,Lawson:1979kx,Dongarra:1990fk} 
and similar
sparse matrix libraries~\cite{Saad:fr}.
%,George:1981uq}. 
Writing programs in
this way promotes readability and maintainability but can be costly in terms
of memory efficiency. Specifically, the retrieval of a large order matrix at
each routine call can have a profound effect on performance even when highly
tuned implementations of the BLAS (e.g.,~\cite{Goto:2006fk})
%(e.g.~\cite{Bilmes:1997ye,Whaley:1998fk,IntelMath:oq,ESSL:kl,Goto:2006fk})
are used.

The authors and other researchers have observed that it can be much more
efficient to call a single specialized routine that performs multiple
operations than to make successive calls to separate BLAS routines (e.g., see
~\cite{gropp01}.
%~\cite{Ashby:uq,Blackford:2002vn,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.
Single routines that carry out more than one linear algebra operation are
known as \emph{composed} BLAS. As an example, consider the pair of
matrix-vector products $q = Ap$, $s = A^Tr$, where $A$ is a matrix and $p$,
$q$, $r$, and $s$ are vectors, that comprise the computational bottlenecks of
the BiConjugate Gradient Method~\cite{Saad:2003fk} and of the
GEMVER kernel examined in Section~\ref{sec:matlab}.  These two operations can
be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can
be rewritten as a composed BLAS consisting of the doubly nested loop
encompassing both matrix-vector products. In the former case, the matrix
$A$ is accessed twice, while in the latter it is accessed only once. Our
preliminary results in prior research and in this paper indicate that 
loop fusion leads to a routine that delivers significantly better performance
than does a pair of calls to the best optimized BLAS {\tt GEMV} routines for
large matrix orders.  Composed routines are the focus of the work presented
here, but much of what we discuss generalizes to a much broader array of
computations.

To generate highly optimized code from a MATLAB prototype of the composed
BLAS operation, we follow a three-step approach, illustrated in
Figure~\ref{fig:process}.  
%
\begin{wrapfigure}{r}{3in}
\vspace{-.2in}
\centering
\includegraphics[width=3in]{figures/process.png}
\caption{The code generation and tuning process.}
\label{fig:process}
\vspace{-.2in}
\end{wrapfigure}
%
To begin, we have developed a compiler that
converts a MATLAB script into simple C code~\cite{Siek}.  After generating
the C code from the high-level MATLAB prototype, we (optionally) use 
the source-to-source automatic parallelization tool PLuTo~\cite{Pluto} %developed at Ohio State University
to optimize for coarse-grained parallelism and locality simultaneously. Using
the results of the PLuTo analysis, we insert annotations into the C code,
which are then processed by our extensible annotation system Orio to generate
many tuned versions of the same operation using different optimization
parameters. Orio then performs an empirical search to select the best among
the multiple optimized code variants.

@@

%\vspace{-.2in}
\subsection{A MATLAB Compiler}
\label{sec:matlab}
%\input{matlab}
In this section we describe the design and operation of the MATLAB to C compiler we have developed~\cite{Siek}.    Figure~\ref{fig:compiler} gives an overview of the compilation process. 
The MATLAB kernel specification is parsed into a high-level intermediate
representation in the form of a dataflow graph, in which each node represents
%
\begin{wrapfigure}{r}{2in}
%\vspace{-.2in}
\centering
\includegraphics[width=2in]{figures/compile.png}
\caption{The MATLAB compiler.}
\label{fig:compiler}
\vspace{-.2in}
\end{wrapfigure}
%
a parameter (e.g., a scalar, matrix, or vector variable) of the kernel or an
operation.  This dataflow graph is then iteratively processed until all of
the implementation choices have been made.  The compilation process consists
of three phases: analysis, refinement, and optimization, which are together
iterated until all of the implementation decisions have been made.  The graph
is then translated into C code.

%\subsubsection{The Dataflow Graph.}
%
%An example dataflow graph for the GEMVER kernel, defined below, is shown
%in Figure~\ref{fig:gemver-dataflow}.
%Each node represents a parameter of the kernel or an operation.  The arrows
%indicate the flow of data. At first the graph specifies what operations are
%to be performed but does not contain any implementation details. The symbol
%$\times$ in the depicted graph stands variously for outer product,
%matrix-vector multiplication, and scalar-vector multiplication, and it does
%not yet specify, for example, whether the outer products compute a row or a
%column at a time of the result matrix.
%\begin{figure}[ht]
%\begin{minipage}{.3\textwidth}
%\begin{align*}
%  A &\gets A + u_1 v_1^T + u_2 v_2^T \\[-0.5ex]
%  x &\gets \beta A^T y + z \\[-0.5ex]
%  w &\gets \alpha A x
%\end{align*} 
%\end{minipage}
%\begin{minipage}{.6\textwidth}
%\includegraphics[width=3in]{figures/gemver-dataflow.png}
%\end{minipage}
%\caption{Dataflow graph (right) for the GEMVER kernel (left).}
%\label{fig:gemver-dataflow}
%\vspace{-.2in}
%\end{figure}

@@

The refinement phase resolves the implementation for each operation node in
the graph into a subgraph defining the details of the chosen algorithm.  Each
subgraph is an abstract representation of the loop that implements the given
operation that also contains an iteration strategy for traversing the
elements of the matrix or vector.
In the optimization step, we apply conditional rewrite rules to optimize the
dataflow graph, for example merging two subgraphs when they share a common
operand. This rule is responsible for fusing the loops of the two
matrix-vector products in the GEMVER kernel.  
The final step performed by the MATLAB compiler when the graph cannot be
refined further is the generation of C code. The generator outputs a C loop
for each subgraph based on a topological sort of the graph.

@@

@@

\begin{figure}[tb]
%\vspace{-.3in}
\begin{center}
\includegraphics[width=.65\textwidth]{figures/orio.png}
\end{center}
\vspace{-.2in}
\caption{Overview of the Orio empirical tuning process.}
\label{fig:orio}
\vspace{-.2in}
\end{figure}

% --------------------------------------------------------------------------

@@

\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.45\textwidth}
\scriptsize
\begin{verbatim}
spec vadd_tune_spec {
 def build { 
  arg build_command = 'mpixlc -O3 -qstrict -lm';
  arg batch_command = 'qsub -n 64 -t 10';
  arg status_command = 'qstat';
  arg num_procs = 64;
 }
 def performance_params { 
  param UF[] = range(1,32);
  param PAR[] = [True, False];
 } 
 def input_params { 
  param N = [10,100,1000,10**4,10**5,10**6,10**7];
 } 
 def input_vars {
  decl int n = N;
  decl double y[N] = 0;
  decl double x1[N] = random;   
  decl double x2[N] = random;   
  decl double x3[N] = random;   
 }
 def search {
  arg algorithm = 'Exhaustive';
 }
}
\end{verbatim}
\end{minipage}\\
\end{tabular}
\vspace{-.1in}
\caption{Orio example: Annotated C source code (left) and tuning specification excerpt for the Blue Gene/P (right).}
\label{fig:orio-example}
%\vspace{-.2in}
\end{figure}

Figure~\ref{fig:orio-example} shows an annotation example used by Orio to
empirically optimize VADD operation on Blue Gene/P. The annotations contain
performance hints that instruct Orio to perform memory alignment
optimization, loop unrolling, and multicore parallelization (using
OpenMP). In addition to these simple optimizations, Orio supports other
transformations such as loop blocking, loop permutation, scalar replacement,
array copy optimization, and some architecture-dependent optimizations. The
right-hand side of Figure~\ref{fig:orio-example} shows separate tuning
specifications used for building and running executable tests, including
performance parameter values, execution environment details, input variable
information, and the search algorithm. Orio also supports parallel search
when parallel resources are available. In this example, the parallel Orio
driver simultaneously executes 64 code variants in the same parallel job. The
commands used by Orio to submit a parallel job and to query its status are
also specified in the tuning specifications. At present, users must create
the tuning specifications manually. When Orio is used in conjunction with
compiler tools, such as the MATLAB compiler described in this paper, it
should eventually be possible to automatically generate the tuning
specifications.

%\vspace{-0.1in}

@@

Table~\ref{tbl:blas-ops} lists the composed BLAS operations used in our
experiments, along with their input and output variables. Vectors are typeset
in lowercase with an overhead arrow. Scalars and matrices are represented as
lowercase and uppercase letters, respectively. A regular uppercase denotes a
row matrix, whereas a bold uppercase symbolizes a column matrix. The extended
MATLAB expression that corresponds to each operation can be seen in the last
column of Table~\ref{tbl:blas-ops}.

\begin{table}[htb] 
%\vspace{-.1in} 
\caption{Composed BLAS operations used in our experiments.}
\label{tbl:blas-ops}
\small
\centering 
%\vspace{-.1in} 
\begin{tabular}{|c|c|c|c|} 
\hline 
Name & Input & Output & Operation  \\ 
\hline 
\hline 
%& & & \\
VADD & $\overrightarrow{w}$,$\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$ &
$\overrightarrow{x} = \overrightarrow{w} + \overrightarrow{y} + \overrightarrow{z}$ \\ [1ex]
\hline 
%& & & \\
ATAX & $A$,$\overrightarrow{x}$ & $\overrightarrow{y}$ & 
$\overrightarrow{y} = A' * (A * \overrightarrow{x})$ \\ [1ex]
\hline 
%& & & \\
 & ${\bf A}$,$a$,$b$, &  & ${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
GEMVER & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & ${\bf B}$,  & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
& $\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\ [1ex]
\hline 
& & & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\ [1ex]
\hline 
\end{tabular} 
\vspace{-.1in} 
\end{table}

@@

\end{tabular}
}
\vspace{-.1in}
\caption{Performance results for different implementations of several composed BLAS operations.}
\label{fig:results}
\vspace{-.2in}
\end{figure}

The performance results (in MFLOP/s) of tuning the VADD operation on the Blue
Gene/P are given in Figure~\ref{fig:results}(a). The ``Base'' label
designates the C implementation generated by the MATLAB compiler. We also
tested the performance of an implementation that calls
\texttt{daxpy} twice using available BLAS libraries.
Finally, we tuned the simple C loop version using Orio, with the performance
annotations previously shown in Figure~\ref{fig:orio-example}. In this
experiment, we measured the performance for both the sequential and parallel
scenarios. We observe that even for this simple operation, the compiler alone
is unable to obtain the same level of performance as the empirically tuned
versions. Furthermore, as expected, the BLAS implementation does not
exploit locality and thus performed worse than the single loop version.
%of the single loop implementation due to l
% however, selecting the correct compiler
%options was crucial for achieving good performance of the base version.

@@

The Xeon performance results (in seconds) of ATAX are shown in
Figure~\ref{fig:results}(b). The Orio-tuned version that incorporates
PLuTo-generated loop fusion optimizations and Orio parallelization
directives achieves the best performance for most problem sizes,
outperforming the Intel MKL version by a factor of 2 to 5.7 and the
compiler-optimized C version by a factor of 4 to 7. The optimizations
performed by both Orio versions include scalar replacement,
vectorization, and loop unroll/jam.
